---
title: "simple-confounding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simple-confounding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A set of assumptions for causal inference

1. Each observation's potential outcomes $Y_i(1)$ and $Y_i(0)$ are independent of other participants' observed exposures
("non-interference"). Similar to typical i.i.d. assumption; fails e.g. for vaccination effects, due to herd immunity.

2. Each observed outcome is equal to the potential outcome corresponding to the observed outcome:
if $X=x$, then $Y = Y(x)$ ("consistency", "no multiple versions of treatment", "well-defined interventions"). 
Can fail if treatment is imprecisely defined; e.g., "effect of heart transplant".

3. The potential outcomes are each statistically independent of the observed treatment, given some vector of observed covariates $Z$: 
$\forall x: Y(x) \perp X |Z$ ("ignorability", "conditional exchangeability", "no uncontrolled confounding"). This assumption holds automatically
if we randomize $X$, which is why randomized controlled trials are considered reliable sources of evidence.

4. The probability of treatment 
The first two assumptions are sometimes jointly referred to as the "Stable Unit Treatment Value Assumption" (SUTVA).

These three assumptions are sufficient for causal inference, but they aren't always necessary! 
However, they are very commonly used, because they are very helpful. Specifically:

Assumption 2 entails that $E[Y(x)|X=x,Z=z] = E[Y|X=x,Z=z]$; thus it allows us to infer 
the conditional expectation of the potential outcome $Y(x)$ from the conditional expectation of the observed outcome $Y$,
for those participants who received $x$.

Assumption 3 entails that $E[Y(x)|X=x,Z=z] = E[Y(x)|Z=z]$; thus it allows us remove $X=x$ from the condition on the right side of 
the previous expression.

Thus, Assumptions 2 and 3 combined give us the following key result:

$$E[Y(x)|Z=z]= E[Y|X=x,Z=z]$$
This result says that if our assumptions are correct, we can accurately estimate 
the conditional distributions of the potential outcomes from the conditional 
distributions of the observed outcomes.

Let's see what we can do with that result.

# Example scenario

Suppose we have a data set of observations for three binary variables, $X$, $Y$, and $Z$.

Suppose we are interested in the effects of $X$ on $Y$, i.e., we are interested in some causal estimands
which are defined as functions of the potential outcomes $Y(x)$.

If we assume that our observed covariate $Z$ satisfies Assumption 3 (ignorability/conditional exchangeability) and that 
Assumptions 1 and 3 also hold, then we have:

$$p(Y(x)=1 | Z=z) = p(Y=1 | X=x, Z=z)$$
And further, by the Law of Total Probability we have:

$$p(Y(x)=1) = \sum_{z\in \{0,1\}} p(Y(x)=1|Z=z)\cdot P(Z=z)$$
$$=\sum_{z\in \{0,1\}} p(Y=1 | X=x, Z=z) \cdot P(Z=z)$$
We can consistently estimate $p(Y=1 | X=x, Z=z)$ and $P(Z=z)$ from the observed joint distribution $p(X,Y,Z)$,
so under our assumptions, we can also consistently estimate $p(Y(x)=1 | Z=z)$ and $p(Y(x)=1)$.

We can also estimate $p(Z=z|X=1)$, which enables us to estimate the Average Treatment effect among the Treated (ATT), $E[Y(1)−Y(0)|X=1]$:

$$E[Y(x)|X=1]= \sum_{z\in \{0,1\}} E[Y(x)|Z=z,X=1]\cdot P(Z=z|X=1)$$
$$= \sum_{z\in \{0,1\}} E[Y(x)|Z=z]\cdot P(Z=z|X=1)$$
$$= \sum_{z\in \{0,1\}} E(Y|X=x,Z=z]\cdot P(Z=z|X=1)$$
Next, let's see how this works in practice.

# Simulation example

We will simulate data from a distribution $p(X,Y,Z)$, specified such that $p(y|x,z) = p(Y(x)=y|Z=z)$.
Specifically, we will construct the data-generating model such that:
$$p(y|x,z) = p(Y(x)=1 | Z=z) = 0.1 + 0.1x + 0.2z + 0.1xz$$
This model entails the following table of conditional potential risks:

```{r, message=FALSE}

library(dplyr) # convenient data-manipulation functions
library(pander) # table formatting functions
library(purrr)
panderOptions("table.split.table", Inf) # make sure wide tables don't get split up

PO_model = tribble(
  ~x, ~z,
  0, 0,
  0, 1,
  1, 0,
  1, 1) |> 
  mutate(`p(Y(x)=1|Z=z)` = 0.1 + 0.1*x + 0.2*z + .1*x*z)

pander(PO_model)
```

We will simulate $Z$ as a Bernoulli random variable with $p(Z=1) = 0.5$,
and we'll simulate $X$ as Bernoulli RV with $p(X=1|Z=z) = .3 + .4z$.
We don't need to assume a causal relationship between $Z$ and $X$; only an association.
We also don't need to assume that $Z$ is a cause of $Y$, either, but due to how we defined
$p(Y(x)=1 | Z=z)$, we have assumed that there is a causal connection between $Z$ and $Y$. 
Let's talk more about the exact nature of that connection later.

Now let's simulate the data; first we'll generate Z:

```{r}

set.seed(1) # control RNG, for reproducibility
n = 10^6 # number of observations in simulated data set; made large to see asymptotic results
data1 = tibble(Z = rbernoulli(n = n, p = 0.5))

```

Next, we'll generate the potential outcomes $Y(0)$ and $Y(1)$. 
We'll do this before simulating $X$, to make it clear that these potential outcomes
don't depend on $X$, given $Z$; thus, our data-generating model satisfies ignorability/conditional exchangeability.

```{r}

data1 = 
  data1 |> 
  mutate(
  `Y(0)` = rbernoulli(n = n, p = .1 + .2*Z),
  `Y(1)` = rbernoulli(n = n, p = .1 + .1*1 + .2*Z + .1*1*Z))

```

Next, we'll simulate $X$, and use $X$ to determine which potential outcome becomes the observed outcome $Y$:

```{r}

data1 = data1 |> 
  mutate(
  X = rbernoulli(n = n, p = .3 + .4*Z),
  Y = if_else(X == 1, `Y(1)`, `Y(0)`)) |> 
  mutate(across(where(is.logical), as.numeric)) # convert from TRUE/FALSE to 0/1 representations, for convenience

```

Since we simulated both potential outcomes for every observation, let's peek behind the curtain
and confirm that our simulation results match our intended potential outcomes model:

```{r}

data1 |> 
  group_by(z = Z) |> 
  summarize(
    `E[Y(0)|Z=z]` = mean(`Y(0)`),
    `E[Y(1)|Z=z]` = mean(`Y(1)`)) |> 
  pander()

```

We can see that these results closely match the previous table. 
Furthermore, we can use the potential outcomes to directly approximate the marginal 
average potential outcomes, $E[Y(0)]$ and $E[Y(1)]$:

```{r}

data1 |> 
  summarize(
    `E[Y(0)]` = mean(`Y(0)`),
    `E[Y(1)]` = mean(`Y(1)`)) |>  
  mutate(
    `E[Y(1) - Y(0)]` = `E[Y(1)]` - `E[Y(0)]` 
  ) |> 
  pander()

```

With a little probability theory, we can work out that the true marginal average 
potential outcomes are indeed $E[Y(0)] = 0.2$ and $E[Y(1)] = 0.35$, and thus $E[Y(1) - Y(0)] = 0.15$.

In practice, we don't observe the complete $Y(0)$ and $Y(1)$ vectors; 
we only observe $Y$ (and $X$ and $Z$). So let's implement the estimation strategies we sketched previously.

Let's start with $E[Y(0)]$ and $E[Y(1)]$. How do we estimate them?

## What not to do:

```{r}

data1 |> 
  summarize(
    `E[Y|X=0]` = mean(Y[X==0]),
    `E[Y|X=1]` = mean(Y[X==1])) |> 
  mutate(
    `E[Y|X=1] - E[Y|X=0]` = `E[Y|X=1]` - `E[Y|X=0]`
    ) |>  
  pander()

```

Comparing this table to the previous one, we can see that $E[Y|X=x] \neq E[Y(x)]$ and 
$E[Y|X=1] - E[Y|X=0] \neq E[Y(1) - Y(0)]$; he apparent risk difference, $E[Y|X=1] - E[Y|X=0]$, is about 10 percentage points larger than the 
true average treatment effect, $E[Y(1) - Y(0)]$.

## Regression analysis

Since we have assumed that $p(Y(x)=1|Z=z) = p(Y=1|X=x,Z=z)$, we might start by estimating $p(Y=1|X=x,Z=z)$
using a regression model. 
Specifically, we can fit a generalized linear model $p(Y=1|X=x,Z=z) = \beta_0+\beta_Xx +\beta_Zz + \beta_{XZ}xz$, like so:

```{r}

glm1 = glm(
  data = data1, 
  formula = Y ~ X * Z, 
  family = binomial(link = "identity"))

glm1 |> summary() |> coef() |> pander()

```

This model is saturated, so a logistic link would produce a numerically equivalent fit compared to the identity link I'm using here;
but the identity link is more convenient to work with because the coefficients correspond to risks and risk differences rather than log-odds and log-odds ratios.

If we compare the estimated coefficients to the potential outcomes model coefficients that used to generate the data, $p(Y(x)=1 | Z=z) = 0.1 + 0.1x + 0.2z + 0.1xz$, we can see that we have approximately recovered the coefficients of that potential outcomes model.

Under our assumptions, we can extract estimates of the conditional potential risks from this regression model; i.e.,

$$\hat E[Y(x)|Z=z] = \hat\beta_0 + \hat\beta_Xx + \hat\beta_Zz + \hat\beta_{XZ}xz$$
```{r}

beta = coef(glm1)

PO_estimates = 
  PO_model |> 
  mutate(
    `p(Y=1|X=x,Z=z)` = beta[1] + beta[2]*x + beta[3]*z + beta[4]*x*z
    # more generally: `p(Y=1|X=x,Z=z)` = predict(glm1, newdata = tibble(X=x,Z=z), type = "response")
    ) 

pander(PO_estimates)

```

By comparing the columns $p(Y(x)=1|Z=z)$ and $\hat p(Y=1|X=x,Z=z)$, we can see that we have succeeded in recovering the underlying causal model.

Now, we can consistently estimate the marginal potential risk $p(Y(x)=1)$ by marginalizing the fitted model over the estimated distribution of $Z$:
$$\hat{p}(Y(x)=1) = \sum_{z\in{0,1}}{\hat{p}(Y(x)=1 | Z=z)}\hat{p}(Z=z)= \sum_{z\in{0,1}}{\hat{p}(Y=1|X=x,Z=z)}\hat{p}(Z=z)$$
This type of estimate is called a "g-computation formula" ("g-formula" for short), and it was popularized by Jamie Robins, although it is actually the same thing as standardization, which has been around for a long time.

### Average Treatment effect on the Treated (ATT)

Regression modeling also makes it relatively simple to compute the ATT. We first need to fit an additional model, $p(Z=z|X=1)$:

```{r}

ATT_model = glm(
  formula = Z ~ X,
  data = data1,
  family = binomial(link = "identity"))

ATT_model |> summary() |> coef() |> pander()

`p(Z=1|X=1)` = predict(ATT_model, newdata = tibble(X=1), type = 'response')

```

(Alternatively, we could estimate $p(Z=z)$ and $p(X=1|Z=z)$ and solve for $p(Z=z|X=1)$ using Bayes' Theorem).

Then the ATT is:

$$P(Y(1)-Y(0)|X=1) = \sum_{z\in(0,1)}{\left[p(Y=1|X=1,Z=z) - p(Y=1|X=0,Z=z)\right] \cdot p(Z=z|X=1)}$$
We can compute the sample-analogue of this quantity like so:

```{r}

PO_estimates |> 
  group_by(z) |> 
  summarize(
    `E[Y(1)-Y(0)|Z=z]` = `p(Y=1|X=x,Z=z)`[x==1] - `p(Y=1|X=x,Z=z)`[x == 0]
  ) |> 
  mutate(
   `p(Z=z|X=1)` = if_else(z == 1, `p(Z=1|X=1)`, 1-`p(Z=1|X=1)`)
  ) |> 
  summarize(
     `E[Y(1)-Y(0)|X=1]` =  sum(`E[Y(1)-Y(0)|Z=z]` * `p(Z=z|X=1)`) 
  ) |> 
  pander()

```

For our particular data-generating model, we know that:


$$P(Y(1)-Y(0)|X=1) = \sum_{z\in(0,1)}{\left[p(Y=1|X=1,Z=z) - p(Y=1|X=0,Z=z)\right] \cdot p(Z=z|X=1)}$$
$$=\sum_{z\in(0,1)}{[(\beta_0+\beta_X+\beta_Zz+\beta_{XZ}z)- (\beta_0+\beta_Zz)] \cdot p(Z=z|X=1)}$$
$$=\sum_{z\in(0,1)}{[\beta_X+\beta_{XZ}z] \cdot p(Z=z|X=1)}$$
$$=\beta_X + \beta_{XZ} \cdot p(Z=1|X=1)$$
$$=\beta_X + \beta_{XZ} \cdot p(Z=1|X=1)$$
$$=0.1 + (0.1 \cdot p(Z=1|X=1))$$
We also have:

$$p(Z=1|X=1) = \frac{p(X=1|Z=1)p(Z=1)}{p(X=1|Z=1)p(Z=1) + p(X=1|Z=0)p(Z=0)}$$
$$=\frac{.7\times.5}{(.7\times.5)+(.3\times.5)}$$
$$=0.7$$
So $P(Y(1)-Y(0)|X=1) = 0.1 + (0.1 * 0.7) = .17$, which mateches our empirical estimate.

## Stratification

Alternatively, if $Z$ is discrete-valued or can be discretized without invalidating ignorability (assumption 3),
and if we have enough data that there are a substantial number in each stratum of $Z$,
then we could stratify the data set by $Z$, 
and compute $E[Y|X=x,Z=z]=E[Y(x)|Z=z]$ directly for each $x,z$ pair: 

```{r}

strata =
  data1 |> 
  group_by(z = Z) |> 
  summarize(
    .groups = "drop",
    `E[Y|X=1,Z=z]` = mean(Y[X==1]),
    `E[Y|X=0,Z=z]` = mean(Y[X==0]),
    `E[Y(1) - Y(0)|Z=z]` = `E[Y|X=1,Z=z]` - `E[Y|X=0,Z=z]`,
    `p(Z=z)` = n()/nrow(data1))

strata |> pander()

```

Here we can see that the stratified means `E[Y|X=x,Z=z]` are approximately equal
to the conditional average potential outcomes, $E[Y(x)|Z=z]$.

We can compute a weighted average of the stratified estimates `E[Y(1) - Y(0)|Z=z]`, 
with weights equal to `p(Z=z)`, to estimate the marginal average treatment effect (ATE), `E[Y(1) - Y(0)]`:

```{r}

`E[Y(1) - Y(0)]` = 
  strata |> 
  summarize(`E[Y(1) - Y(0)]` = sum(`E[Y(1) - Y(0)|Z=z]` * `p(Z=z)`)) |> 
  pander()


```

In this example, regression and stratification are actually mathematically equivalent. However, if $Z$ were continuous, then stratification would require choosing a discretization of $Z$ that we think is sufficiently fine-grained, whereas we could perform the regression analysis using the observed continuous Z and replacing the summation step $$\hat{p}(Y(x)=1) = \sum_{z\in{0,1}}{\hat{p}(Y(x)=1 | Z=z)}\hat{p}(Z=z)= \sum_{z\in{0,1}}{\hat{p}(Y=1|X=x,Z=z)}\hat{p}(Z=z)$$ with an integration step
$$\hat{p}(Y(x)=1) = \int_{z\in \mathbb{R}}{\hat{p}(Y(x)=1 | Z=z)}\hat{p}(Z=z)dz= \int_{z\in \mathbb{R}}{\hat{p}(Y=1|X=x,Z=z)}\hat{p}(Z=z)dz$$

We also don't need to use a saturated regression model; for example, if we are confident that the interaction term is unnecessary ($\beta_{XZ}=0$), we could state that assumption and remove it from the model. Then the regression result would not match the stratification result; the regression approach would lose flexibility and gain precision.

## Propensity scores

Sometimes, it may be easier to fit a model for $p(X=x|Z=z)$ than a model for $p(Y=y|X=x,Z=z)$, for example if the functional form of $p(X=x|Z=z)$ is simpler or better-understood from prior research. In such cases, we can use $p(X=x|Z=z)$, which we call the "propensity score" (i.e., propensity of treatment score), to estimate causal effects. There are several ways to use propensity scores, including regression adjustment, matching, and weighting. Here I'll demonstrate weighting adjustment: we can estimate $p(X=x|Z=z)$ and then regress $Y$ on $X$ with weights equal to $1/\hat p(X=x|Z=z)$:

```{r}

PS_model = glm(
  data = data1,
  family = binomial, # model is saturated, so link function doesn't matter
  X ~ Z)

data1 =
  data1 |> 
  mutate(PS = predict(PS_model, newdata = tibble(Z), type = "response"))

glm4 = glm(
  data = data1,
  family = quasibinomial(link = "identity"), # need quasibinomial for weights
  Y ~ X,
  weights = 1/PS
)

glm4 |> summary() |>  coef() |>  pander()


```

```{r}

PS_strata =
  data1 |> 
  group_by(PS) |> 
  summarize(
    .groups = "drop",
    `E[Y|X=1,PS=s]` = mean(Y[X==1]),
    `E[Y|X=0,PS=s]` = mean(Y[X==0]),
    `E[Y(1) - Y(0)|PS=s]` = `E[Y|X=1,PS=s]` - `E[Y|X=0,PS=s]`,
    `p(PS=s)` = n()/nrow(data1))

PS_strata |> pander()

```

Here we can see that the conditional means `E[Y|X=x,Z=z]` are approximately equal to the 
conditional potential outcomes, $E[Y(x)|Z=z]$.

Analogously to when we stratified on $Z$, we can average the stratum-specific effect estimates, 
weighting by the distribution of the propensity score strata, to estimate the marginal 
average treatment effect `E[Y(1) - Y(0)]`:

```{r}

`E[Y(1) - Y(0)]` = 
  PS_strata |> 
  summarize(`E[Y(1) - Y(0)]` = sum(`E[Y(1) - Y(0)|PS=s]` * `p(PS=s)`)) |> 
  pander()


```

The result is again very close to the underlying "true" ATE derived by our data-generating model, $E[Y(1) - Y(0)] = 0.15$.
